---
title: "Randomized complete block design"
date: 2022-10-30
abstract: "One-way ANOVA & pairwise comparison post hoc tests in a randomized complete block design."
---

```{r}
#| include: false
source(here::here("helpersrc.R"))
```

```{r}
# (install &) load packages
pacman::p_load(
  broom,
  conflicted,
  desplot,
  emmeans,
  ggtext,
  multcomp,
  multcompView,
  tidyverse)

# handle function conflicts
conflict_prefer("filter", "dplyr") 
conflict_prefer("select", "dplyr")
```

# Data

This example is taken from Chapter *"2 Randomized complete block design"* of the course material "Mixed models for metric data (3402-451)" by [Prof. Dr. Hans-Peter Piepho](https://www.uni-hohenheim.de/organisation?tx_base_lsfcontentadmin%5BlsfPerson%5D=6257). It considers data published in Clewer and Scarisbrick [-@clewer2001] from a yield (t/ha) trial laid out as a randomized complete `block` design (3 blocks) with `cultivar` (4 cultivars) being the only treatment factor. Thus, we have a total of 12 plots.


